---
title: "simplaceUtil"
author: "Gunther Krauss"
date: "`r Sys.setlocale('LC_ALL','English.UTF-8');format(Sys.Date(),format='%B %d,
  %Y')`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    fig-width: 10
    fig-height: 8
    toc: yes
    df_print: kable
  pdf_document:
    number_sections: yes
    df_print: kable
    toc: yes
vignette: >
  %\VignetteIndexEntry{simplaceUtil} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE, error=FALSE}
# withSimplace <- FALSE
# try({
#   withSimplace <- !is.na(simplace::findFirstSimplaceInstallation())
# })

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

includeGraphic <- function(gr,name) {
  
    type <- ifelse(knitr::is_latex_output(),'.pdf','.png')
    fn <- tempfile(fileext = type)
    DiagrammeR::export_graph(gr, fn, width = 2000)
    fn <- "a"
    if(!file.exists(fn))
    {
      fn <- system.file("img",paste0(name,type),package="simplaceUtil")
    }
    try(knitr::include_graphics(fn), silent=TRUE)
}


```

# Exploring solutions

## Fetch information from solution


```{r}
library(simplaceUtil)
```

We can read a solution xml file via `getSolutionFromFile` and get its content as an `xml_document` object. From the xml object we can get the components, user variables and links between components as dataframes.

```{r}
solfile <- system.file("solution","Complete.sol.xml",
                       package="simplaceUtil")

sol <- getSolutionFromFile(solfile)
components <- getComponents(sol)
variables <- getUserVariables(sol)
links <- getLinks(sol, components)
```


As a shorthand we can use `getElementsFromSolutionFile`, which returns the xml object as well as the dataframes in a list.

```{r}
el <- getElementsFromSolutionFile(solfile)
sol <- el$solution
components <- el$components
variables <- el$variables
links <- el$links
```

The components dataframe contains not only sim components, but also interfaces, resources and outputs.
```{r, echo=FALSE}
components$ref <- gsub('$', '\\$', components$ref, fixed = TRUE)
```

```{r}
components[c(1:3,30:31,42:44),c("id","type","subtype","nr")]
```


The links dataframe contains the links between resources, simcomponents, outputs, but also dependencies via key variables or via variables used in rules. Relationships between resources and transformers are included too.

```{r}
links[c(1:2,23,300,313,322,325),]
```


User variables show the values as defined in the solution. Notice that user variables can be changed when running a solution in project mode.

```{r}
variables[c(1:5,18:20),]
```

```{r}
head(components[,c("id","ref")])
```


```{r, echo=FALSE}
components$ref <- gsub('\\$', '$', components$ref, fixed = TRUE)
```


```{r}
components$ref <- replaceVariablesWithValues(
  components$ref,
  variables, 
  additional=c("_WORKDIR_"="~"))
head(components[,c("id","ref")])
```




## Working with component and links dataframe

Subsetting the components dataframe to get only normal components

```{r}
ncomp <- components[components$type=="simcomponent" & components$subtype=="normal",]
ncomp$id
```

Getting only links between normal components

```{r}
nlinks <- links[links$from %in% ncomp$id & links$to %in% ncomp$id,]
head(nlinks)
```

Counting the links between components

```{r}
nlinks <- aggregate(name ~ from + to , data=nlinks, length)
nlinks <- nlinks[order(-nlinks$name),]
head(nlinks,10)
```





## Get Metadata about solution

`getSolutionInfoAsDataframe` gives information about the solutions content 
(simcomponents) as well as the solution itself (filename, modification date and
possibly simulation experiment information derived from folder structure).


```{r}
wd <- system.file(package="simplaceUtil")
solfiles <- list.files(wd,pattern=".sol.xml",recursive=TRUE,full.names = TRUE)

df <- do.call(rbind,
        lapply(solfiles,getSolutionInfoAsDataframe,workdir=wd))
df$file <- basename(df$file)
df[df$type=="simcomponent",c("id","file")] 
```


# Visualising solution structure

A graph showing components and links can be created from a solution file:

```{r}
graph <- solutionToGraph(solfile)
```
```{r eval=FALSE}
DiagrammeR::render_graph(graph)
```
```{r echo=FALSE, out.width="100%"}
includeGraphic(graph,"graph")
```

We can restrict the graph to specific component types and the neighborhood of
selected components:

```{r}
subgraph <- getNeighborhood(graph, c("SlimWater","SlimRoots"),
                            distance=1, type="simcomponent")
```
```{r eval=FALSE}
DiagrammeR::render_graph(subgraph)
```
```{r echo=FALSE, out.width="100%"}
includeGraphic(subgraph,"subgraph")
```


# Modifying solutions

A solution can be modified on the fly to meet requirements for sensitivity analysis or calibration  by

- removing non-memory outputs for performance reasons
- adding memory outputs with target variables
- adding user variables for parameters that will be varied
- replacing variables from resource by user variables


\footnotesize
```{r}
solution_candidate <- system.file("solution","Yield.sol.xml",
                       package="simplaceUtil")
sol <- getSolutionFromFile(solution_candidate)
newsol <- sol |>
  removeNonMemoryOutputs() |>
  addMemoryOutput("Yields",frequence = "YEARLY") |>
  addOutputVariable("Yields","Year","CURRENT.YEAR","INT") |>
  addOutputVariable("Yields", "ClimateZone","vClimateZone","CHAR") |>
  addOutputVariable("Yields", "Yield","LintulBiomass.sWSO","DOUBLE","MAX") |>
  addUserVariable("vLUE",3.0,"DOUBLE") |>
  addUserVariable("vSLA",0.02,"DOUBLE") |>
  addUserVariable("vRGRL",0.009,"DOUBLE") |>
  replaceVariable("crop.LUE","vLUE") |>
  replaceVariable("crop.SLA","vSLA") |>
  replaceVariable("crop.RGRL","vRGRL")

soltext <- getTextFromSolution(newsol)

cat(paste(substr(soltext,1,1500),"..."))
```

The solution can be saved as xml file:

```{r eval=FALSE}
solution_modified <- "simulation/solution/YieldModified.sol.xml"  
writeSolutionToFile(newsol, solution_modified) 
```


# Standard transformation and visualisation of outputs

With `parseDate` the standard Date column of Simplace output files is converted to Date.

```{r}
outfile <- system.file("output","water.csv",
                       package="simplaceUtil")
water <- read.csv(outfile, header=TRUE, sep=";")
water <- parseDate(water)
```


Multiple scalar values can be plotted in a common diagram.

```{r fig.width=7, fig.height=5}
plotScalarOutput(water,"CURRENT.DATE",c("Evaporation","Transpiration"))
```

Default plot for layered values.

```{r fig.width=7, fig.height=5}
plotLayeredOutput(water,"RetainedWater")
```


Transform dataframe from wide format to long format.

```{r}
water_long <- transformLayeredData(water)
head(water_long[,c(2,9:11)])
```


# Simple GUI as Shiny App

```{r eval=FALSE}
runSimplaceGuiApp()
```

